Template for indicator documentation

### 2023-09-14 slitasje.Rmd
Add some cross refs and hyperlinks to GH issues

# Anthropogenic disturbance to soils and vegetation ADSV {#slitasje}

@@

1.  Import [Nature type data](#NTM) data set (incl. GRUK) and [ANO data
    set](#ANO-import)

@@

4.  For each polygon, extract the NiN-variable that has the lowest value
    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype).

5.  [Combine](#combine-nt-ano) data sets

@@

9.  Confirm relationship between infrastructure index and indicator
    values to justify the extrapolation

@@

The data goes back to 2018. I therefore bulk all the data from 2018 to 2022 into one time step. I then use the mean date for the raw data,
and define the variable as belonging to the year 2020 (read more
[here](#scaled-slitasje-variable)).

@@

The indicator is tagged to the *Økologisk egenskap* called
**Primærproduksjon** (Primary productivity). This is tentative, and
perhaps *abiotiske forhold* is more suited. But the thought behind the
choice is that *slitasje* affect the potential for primary productivity.

@@

_Slitasje_ is not thought to exhibit collinearity with any other indicator at
the present.

@@

The reference condition is one with minimal negative human impact. This
is also true for semi-natural ecosystems. For example, in the reference
condition, *slitasje* in semi-natural ecosystems is defined to be zero,
even though at no point in time would this condition be realized.

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values

@@

The upper reference value is 0 (no *slitasje*), and the lower reference
value is 100%. Note that 100% _slitasje_ does not mean that all the area
must be scarred or damaged, but that all hypothetical quadrats around a
point is affected (the variable is frequency-driven, and not
amount-driven). The threshold for good ecosystem condition is 10%
damage. Perhaps is should be set somewhat higher, like 20 or 30%.
The indicator value could be interpretted differently at the polygon
vs at the landscape scale. At the polygon scale, 10% damage may seem 
like not that much, whilst at the landscape scale, 10% seems more serious.
This final threshold value should be debated and illustrated with examples.

Read about the normalisation [here](##scaled-slitasje-variable).

@@

```{r eight-step, fig.cap="Eight step condition scale"}
knitr::include_graphics("images/mdirPRscale.PNG")
```

@@

I here transfer these ordinal variables over to a continuous scale.

@@

First I will combine 7TK and PRTK, and also 7SE and PRSL

@@

This variable is also recorded in
[GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
The nature type data set I'm working on here includes this data already
(presently only 2021 included). GRUK also records a related variable:
`% cover in 5m radii circles`, which is much more detailed. This data is
not published. In any case it is better to use the harmonized data set
in our case.

@@

Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
placed locations each with 18 sample points. In each sample point a
circle of 250 m2 is visualised, and the main ecosystem is recorded.
Depending on the main ecosystem, certain NiN variables are also
recorded. 7SE is recorded for våtmark, but not semi-natural areas or
naturally open areas. 7TK is recorded in våtmark and naturally open
areas only.

@@

This data set only contains data from year 2019 and 2021. We need to
update this data set later. It is not clear why there is no data from
2020.

@@

Combining these two variables 7TJ and 7SE into one, same as for the nature
type data.

@@

#### Validate

>>>>

# Anthropogenic disturbance to soils and vegetation (ADSV) {#slitasje}

@@

1.  Import [Nature type data](#NTM) data set and [ANO data
    set](#ANO-import)

@@

4.  For each polygon, extract the variable that has the lowest value
    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype). The four variables are 7TK, 7SE (both NiN2-variables) and PRTK and PRSL (both defined by the Norwegain Environmental Agency)

5.  [Combine](#combine-nt-ano) data sets (nature types and ANO) but keep ecosystems separate.

@@

9.  Confirm relationship between infrastructure index and indicator
    values to [justify the extrapolation](#validate-slitasje)

@@

The data goes back to 2018. I therefore [bulk all the data from 2018 to 2022](https://github.com/NINAnor/ecosystemCondition/issues/101#issuecomment-1719170208) into one time step. I then use the mean date for the raw data,
and define the variable as belonging to the year 2020 (read more
[here](#scaled-slitasje-variable)).

@@

The indicator is tagged to the *Økologisk egenskap* called
**Primærproduksjon** (Primary productivity). This is [somewhat tentative](https://github.com/NINAnor/ecosystemCondition/issues/96), and
perhaps *abiotiske forhold* is more suited. But the thought behind the
choice is that *ADSV* affect the potential for primary productivity.

@@

_ADSV_ is not thought to exhibit collinearity with any other indicator at
the present.

@@

The reference condition is one with minimal negative human impact. This
is also true for semi-natural ecosystems. For example, in the reference
condition, *ADSV* in semi-natural ecosystems is defined to be zero,
even though at no point in time would this condition be realized.

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
For a further discussion on this topic, see [GitHub issue 88](https://github.com/NINAnor/ecosystemCondition/issues/88).

@@

The upper reference value is 0% (no *damage*), and the lower reference
value is 100%. Note that 100% _damage_ does not mean that all the area
must be scarred or damaged, but that all hypothetical quadrats around a
point is affected (the variable is frequency-driven, and not
amount-driven). The threshold for good ecosystem condition is set to 10%
damage.

The indicator value could be interpreted differently at the polygon
_vs_ landscape scale. At the polygon scale, 10% damage may seem 
like not that much, whilst at the landscape scale, 10% seems more serious.

Read about the normalisation [here](#scaled-slitasje-variable).

@@

```{r eight-step, fig.cap="Eight step condition scale"}
knitr::include_graphics("images/ninAR8.PNG")
```

@@

I here transfer these ordinal variables over to a continuous scale based on the mean value of each step, but we acknowledge that there are important assumptions being made here, for example that the mean _real_ value for 7TK or 7SE 3 is the mid-point in the range from 0.5 to 1 (i.e. 0.75). This is probably not correct. The large range of the steps, particularly for the 7SE and 7TK variables, poses a problem for assessing ecological condition at the polygon scale, but less so at aggregated scales. Hence, the indicator should not be presented and interpreted at the polygon scale.

@@

First I will combine 7TK and PRTK, and also 7SE and PRSL using the one-out-all-out principle. This is a more conservative approach than for example simply adding these together, but we do it this way because the two variables are likely to overlap considerably at small spatial scales.

@@

This variable is also recorded in
[GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
This is a stratified and intensive monitoring program for calcareous naturally open areas around the Oslo fjord. 
The nature type data set I'm working on here includes some data from GRUK already
(presently only 2021 included). The full GRUK data set also contains a related variable:
`% cover in 5m radii circles`, which is much more detailed than the four variable we have looked at so far. This data is
not published. In any case it might be better to use the harmonized data set
in our case.

@@

Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
placed locations each with 18 sample points. In each sample point a
circle of 250 m2 is visualised, and the main ecosystem is recorded.
Depending on the main ecosystem, certain NiN variables are also
recorded. 7SE is recorded for våtmark (wetlands), but not semi-natural areas or
naturally open areas. 7TK is recorded in wetlands and naturally open
areas only.

@@

This data set only contains data from year 2019 and 2021. We need to
update this data set later. It is not clear why there is no data from
2020. This topic is maintained as a [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/101).

@@

Combining these two variables 7TK and 7SE into one, same as for the nature
type data.

@@

#### Validate {#validate-slitasje}
